\documentclass[10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{parskip}
\usepackage[a4paper, margin=1.5in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{array}

\renewcommand{\arraystretch}{1.5}

\title{Key-Value DB Feasability Study and Implementation}
\date{11/11/2019}
\author{Federico Fregosi, Mirko Laruina,\\
        Riccardo Mancini, Gianmarco Petrelli}

\begin{document}
\pagenumbering{gobble}
\maketitle
\vfill
\setcounter{tocdepth}{2}
\tableofcontents
\vfill
\clearpage
\setcounter{page}{1}
\pagenumbering{arabic}

\section{Feasability study}

Since there are few entities in our database, we decided to implement the whole 
database as a key-value database. By doing so, we learned how to manage different
kinds of relationships between entities in a key-value database.

In the conversion from a relational database to a key-value one, we decided to 
proceed as follows:
\begin{enumerate}
    \item identify a naive design for the keys
    \item check the design again the use-cases to check problems or 
        inefficiencies
    \item add additional keys in order to fix the problems or speed up some use-cases
    \item merge all changes in the final design
\end{enumerate}

\subsection{Naive design}
We decided to generate ids incrementally by saving in the database the next id
to be used. This value will be updated each time a new element is generated.

Another design choice we made is to represent the many to many relationship 
between users and chats through a list of chat members within the chat inside 
a structured attribute of the chat entity. 
Furthermore, we'll later see that this is not sufficient in order to provide 
efficient read operations and the redoundant list of chats of each user will 
be added as a structured attribute of the user entity.

Finally, the chat messages are saved under the chats and have an id that is 
unique only in the chat. Since messages cannot be deleted, no list is necessary.

The naive conversion is shown in the next table. A word preceeded by a dollar 
(\$) indicates a value (that is the id in every case).

\begin{center}
\begin{tabular}{ | c | c | }
    \hline
    \textbf{Key} & \textbf{Value} \\\hline
    user:\$userId:username & Users.username \\\hline
    user:\$userId:password & Users.password \\\hline
    session:\$sessionId:expiry & Sessions.expiry \\\hline
    session:\$sessionId:userId & Sessions.userId \\\hline
    users:nextId & Next id for a user \\\hline
    chat:\$chatId:name & Chats.name \\\hline
    chat:\$chatId:lastActivity & Chats.lastActivity \\\hline
    chat:\$chatId:admin & Chats.adminId\\\hline
    chat:\$chatId:messages:nextId & Next id for a chat \\\hline
    chat:\$chatId:message:\$messageId:text & Messages.text \\\hline
    chat:\$chatId:message:\$messageId:timestamp & Messages.timestamp \\\hline
    chat:\$chatId:message:\$messageId:sender & Messages.senderUserId \\\hline
    chats:nextId & Next id for a chat \\\hline
    chatmembers:\$chatId:\$userId & Empty. Present iff user is member of chat \\\hline
\end{tabular}
\end{center}

\subsection{Use-cases analysis}
Since mapping between use-cases and DB-abstraction-layer methods is 
straight-forward, this analysis will take into consideration the 
methods of the database abstraction layer interface. 

Let's first analyze the methods that give some problems:

\subsubsection{getChats}

\begin{lstlisting}[language = Java]
/**
* Returns the list of chats for the user identified by 
* the given userId.
*
* @return the list of chats or null in case of error.
*/
List<Chat> getChats(long userId);
\end{lstlisting}

The above methods is inefficient because, in order to retrieve the chat a user 
is member of, I should try all chat ids to see if the key exists. As an alternative, 
I could take advantage of the ordering property of LevelDB but the same problem 
would arise in the complimentary method that retrieves the chat members.

Since the number of chats a user is member of and the number of chat members 
are small in practice we decided to replace the intermediate entity with two 
structured attributes of chat and user, respectively (see the table below). 
The attributes hold a 
comma-separated list of ids. The redoundancy is necessary in order to make 
both operations of retrieval of chat members efficient. 

\begin{center}
    \begin{tabular}{ | c | c | }
        \hline
        chat:\$chatId:members & List of chat members \\\hline
        user:\$userId:chats & List of chat ids where the user is present \\\hline
    \end{tabular}
\end{center}

This implementation increases the complexity and the time required for all those
other methods that modify said lists, namely:
\begin{lstlisting}[language = Java]
    boolean addChatMember(long chatId, long userId);
    boolean removeChatMember(long chatId, long userId);
\end{lstlisting}

This is due to the fact that both lists must be downloaded, edited and uploaded back.
However, this is not a big deal since addition and removal are fairly unfrequent 
operations.

Another operation that is involved by this change is \textit{checkChatMember} (see 
below). Compared to the ``naive'' implementation, it is slower but the time penalty
is not as significant as in the previous case, due to the fact that a chat holds 
few members in practice.

\begin{lstlisting}[language = Java]
    boolean checkChatMember(long chatId, long userId);
\end{lstlisting}

\subsubsection{getUser}

\begin{lstlisting}[language = Java]
/**
* Returns user identified by the given username.
*
* @return the user in case of success, null otherwise.
*/
User getUser(String username);
\end{lstlisting}

This method is inefficient using the proposed naive implementation since it 
would require a linear search among all users. 
The use of a reverse index to map usernames to ids would make it more efficient, i.e.:

\begin{center}
    \begin{tabular}{ | c | c | }
        \hline
        username:\$username:userId & Users.userId \\\hline
    \end{tabular}
\end{center}

\subsubsection{existsChat}

\begin{lstlisting}[language = Java]
/**
* Returns true if there exists a private chat between user1 and 
* user2.
*
* NB: a private chat is a chat with only two members.
*
* @return {@code true} if it exists, 
*         {@code false} otherwise or in case of errors.
*/
boolean existsChat(long user1, long user2);
\end{lstlisting}

This method can be implemented in a simple way iterating through the 
user's chats and its members. A more efficient way would be inserting a
redundant list of private chats the user is involved in.
However, since this is just a proof-of-concept implementation and this method gets 
called only on a not frequent operation (chat creation), we decided not to
perform this optimization.

\subsubsection{getChatMessages}

\begin{lstlisting}[language = Java]
/**
* Returns a list of messages for the given chat, in the given 
* time range, up to the given number of elements, sorted in 
* ascending message sending time.
*
* @param chatId id of the chat whose messages are to be retrieved
* @param from start of the time range (included). It can be null, 
*        whose meaning is that there is no lower bound.
* @param to end of the time range (excluded). It can be null, 
*        whose meaning is that there is no upper bound.
* @param n maximum number of elements to return. 
*        If from is not null, messages are counted from 
*        {@code from} up to n or {@code to}. If from is null, 
*        messages are counted from {@code to} up to n or {@code from}.
* @return the list of messages or null in case of error.
*/
List<Message> getChatMessages(long chatId, long from, long to, int n);
\end{lstlisting}

Even though it may look tricky, this query is in fact quite trivial since 
message ids are strictly monotone and adjacent, due to the id generation 
strategy we imposed and the impossibility to delete messages.

To tell the truth, this API was originally thought to take a time range, making its 
implementation more complicated using a key-value database. However, during the
first design phases we decided to change it in order to make its use more 
flexible among different database tecnologies.

Another problem with this API was its slowness due to the three
operations required to read the message from the database (there are 
3 fields to be fetched). This problem was highlighted by the benchmarks
of section~\ref{sec:bench}. The solution is to aggregate the three attributes 
in a single structured attribute since they are always fetched together.

\subsubsection{Trivial methods}
Let's finally list the 
methods whose implementation is trivial and presents no issues using the
proposed structure\footnote{If you're interested in their implementation, you 
can have a look at the code in the \emph{LevelDBAdapter} class.}. 

\begin{lstlisting}[language = Java]
long addChatMessage(Message message);
long createChat(String name, long adminId, List<Long> userIds);
boolean deleteChat(long chatId);
Chat getChat(long chatId);
long createUser(User user);
long getUserFromSession(String sessionId);
boolean setUserSession(UserSession user);
boolean removeSession(String sessionId);
\end{lstlisting}

\subsection{Final design}
\begin{center}
\begin{tabular}{ | c | c | }
    \hline
    \textbf{Key} & \textbf{Value} \\\hline
    user:\$userId:username & Users.username \\\hline
    user:\$userId:password & Users.password \\\hline
    user:\$userId:chats & List of chat ids where the user is present \\\hline
    username:\$username:userId & Users.userId \\\hline
    session:\$sessionId:expiry & Sessions.expiry \\\hline
    session:\$sessionId:userId & Sessions.userId \\\hline
    users:nextId & Next id for a user \\\hline
    chat:\$chatId:name & Chats.name \\\hline
    chat:\$chatId:lastActivity & Chats.lastActivity \\\hline
    chat:\$chatId:members & List of chat members \\\hline
    chat:\$chatId:admin & Chats.adminId\\\hline
    chat:\$chatId:messages:nextId & Next id for a chat \\\hline
    chat:\$chatId:message:\$messageId & Messages.* \\\hline
    chats:nextId & Next id for a chat \\\hline
\end{tabular}
\end{center}

\section{Implementation}
The Java implementation has been done using \emph{LevelDB}\footnote{More information
at: \url{https://github.com/google/leveldb}}. It is a simple Key-Value database, 
providing just three methods to access the store: \emph{put}, \emph{get} and 
\emph{delete}.

Since we included a Database-abstraction layer in our application server, the 
implementation required just adding a new Java class implementing the 
\emph{DatabaseAdapter} interface.

\section{Testing and Evaluation}

The \emph{Key-Value} implementation went under the same testing methods as the 
JPA one (unit tests and final integration test). Furthermore, since our Java 
backend provides a convenient HTTP API, we were able to use 
the open-source software Siege\footnote{More information at: 
\url{https://www.joedog.org/siege-home/}} in order to benchmark the different 
database backends as described in the following section. This was also a way
to test the robustness of our implementation.

\subsection{Benchmarking}
\label{sec:bench}

The benchmark was carried out using Siege. We simulated one user sending multiple
requests and we measured the average \emph{transactions per second} (tps) for 
5000 requests over three runs.

\paragraph{Disclaimer} 
The tests were not done in a rigorous way but they were executed just to get an 
idea on the capabilities of the different database technologies. 

\begin{figure}[h!]
    \centering
    \begin{tabular}{ | c | c | c | c |}
        \hline
        \textbf{Database/Implementation} & get chat list & send messages & get messages \\\hline
        SQL/plain & 397 (309) & 217 & 13.05 \\\hline
        SQL/JPA & 278 (225) & 136 & 9.69 \\\hline
        LevelDB/no\_aggregation & 1022 (587) & 1050 & 3.53 \\\hline
        LevelDB/aggregated & 1033 (527) & 950 & 12.81 \\\hline
    \end{tabular}
\end{figure}

In the first column, the value between parenthesis represents the value 
for the first run, which is the minimum and has been exluded from the average, 
since it is the only one that didn't take advantage of caching (as the other two runs).

From the results it can be seen that:
\begin{itemize}
    \item plain SQL is more efficient than JPA.
    \item Message field aggregation in LevelDB leads to more than 3x better performances.
    \item LevelDB is much faster than SQL in the first two operations, while shows
            similar performances in reading many values (as in the last operation).
\end{itemize}

We also tried increasing the number of concurrent users, obtaining an overall 
better tps (obviously), but showing the same ratios between different backends.

\end{document}
