\documentclass[titlepage, 12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{parskip}

\usepackage[a4paper, margin=1.5in]{geometry}

\title{Task 0 Specification}
\date{08/10/2019}
\author{Federico Fregosi, Mirko Laruina, \\Riccardo Mancini, Gianmarco Petrelli}


\begin{document}
\maketitle

\section{Application overview}
The application is a messaging system where registered users can create an 
account, exchange text messages and make groups.

A registered user can initiate a chat with another user, create a new group chat
(of which he becomes the admin) and send messages to the chats he belongs to,
as well as receiving messages from those chats. He can also leave a group.

A group admin can add and remove new users to the group. He cannot assign his
powers to another user in the group and if he leaves the group, the latter 
is deleted.

Everytime a user views a chat, all the latest messages from the chat are 
fetched from the server and shown to the user.

\section{Actors}
Anonymous user, registered user, group admin and a time-based event.

\section{Functional specifications}
An \textbf{anonymous user} must be able to register in order to become a 
\emph{registered user}.

A \textbf{registered user} must be able to:
\begin{itemize}
    \item Send a message to a chat
    \item Read chat messages
    \item Create a private chat
    \item Create a group chat
\end{itemize}

A \textbf{group admin} must be able to:
\begin{itemize}
    \item Add users to the group
    \item Remove users from the group
\end{itemize}

The \textbf{time-based event} updates the user interface on regular intervals
to show new received messages from the current chat, if any.

\section{Software Architecture}
The proposed software architecture is a classic three-layer architecture:

\begin{enumerate}
    \item database (MySQL)
    \item server back-end (Java + Spring)
    \item user web app (ReactJS)
\end{enumerate}

\end{document}
